SILVR: Conditional diffusion model for molecule generation without additional training

### 2023-03-15 paper/manuscript.tex

Considering all results with respect to the control EDM (rs=0.0), it can be seen that at intermediate values of rs the SILVR protocol produces molecules of equal quality to that of the unmodified EDM, while also being conditionally generated. We therefore claim that if a diffusion model can be successfully trained to produces random high quality drug like structures, SILVR should be applicable to guide generation towards desired regions of chemical space without harming the quality of molecules. While our method poses a direct interface between crystallographic fragment data and de novo molecular generation, we believe future work on SILVR could be expanded further. To answer structure based drug design, a scoring function of protein surface potentials could be used to guide atoms towards energetically favoured pocket regions; issues of synthetic accessibility might be solved by clever design of refinement methods; and interfacing with alternative data sources (such as NMR) might be achievable.

@@

The combined set of case studies demonstrates the ability of SILVR to complete both fragment merging and linking type problems, without any a priori knowledge of these design challenges. It should be noted however that the samples generated by SILVR are often of poorer quality than the samples selected in Figure X. Across all samples around half were determined by RDKit to be fragmented, meaning the sample contained two or more distinct molecular graphs. It was observed qualitatively that fragmented samples typically contained corrupted structures (multiple fragmentations, linear carbon chains, flattened rings, etc). We believe this fragmentation is triggered during intermediate steps of denoising, resulting in an unstable latent representation and subsequently poor EDM inference. Fragmentation becomes of particular issue for linker design type SILVR experiments (figure XE and F), where the reference coordinates direct the latent variables away from eachother, triggering fragmentation. For these experiments 65\% of all samples were fragmented. Further work is needed both with EDM and with SILVR to reduce these rates of fragmentation. Additionally, as is common for generative models, the synthetic accessibility of most sampled molecules is very poor. For our experiments the synthetic accessibility of SILVR generated molecules resembles the performance of the unmodified EDM, so in order to achieve synthetically accessible samples with SILVR an improved EDM will need to be designed.

>>>>

Considering all results with respect to the control EDM (rs=0.0), it can be seen that at intermediate values of rs the SILVR protocol produces molecules of equal quality to that of the unmodified EDM, while also being conditionally generated. We, therefore, claim that if a diffusion model can be successfully trained to produce random high-quality drug like structures, SILVR should be applicable to guide generation towards desired regions of chemical space without harming the quality of molecules. While our method poses a direct interface between crystallographic fragment data and \textit{de novo} molecular generation, we believe future work on SILVR could be expanded further. To answer structure-based drug design, a scoring function of protein surface potentials could be used to guide atoms towards energetically favoured pocket regions; issues of synthetic accessibility might be solved by clever design of refinement methods; and interfacing with alternative data sources (such as NMR) might be achievable.

@@

The combined set of case studies demonstrates the ability of SILVR to complete both fragments merging and linking type problems, without any a priori knowledge of these design challenges. It should be noted however that the samples generated by SILVR are often of poorer quality than the samples selected in Figure X. Across all samples around half was determined by RDKit to be fragmented, meaning the sample contained two or more distinct molecular graphs. It was observed qualitatively that fragmented samples typically contained corrupted structures (multiple fragmentations, linear carbon chains, flattened rings, etc). We believe this fragmentation is triggered during intermediate steps of denoising, resulting in an unstable latent representation and subsequently poor EDM inference. Fragmentation becomes of particular issue for linker design type SILVR experiments (figure XE and F), where the reference coordinates direct the latent variables away from each other, triggering fragmentation. For these experiments, 65\% of all samples were fragmented. Further work is needed both with EDM and with SILVR to reduce these rates of fragmentation. Additionally, as is common for generative models, the synthetic accessibility of most sampled molecules is very poor. For our experiments, the synthetic accessibility of SILVR-generated molecules resembles the performance of the unmodified EDM, so in order to achieve synthetically accessible samples with SILVR an improved EDM will need to be designed.
